CSE530 Project Final Report\\ Bandwidth-Aware Memory Hierarchy Design with Hybrid Memory Technologies

### 2010-11-17 trunk/paper.tex
paper.tex
Re-organize the paper structure and initialize some sections

\documentclass[12pt,letterpaper, singlecolumn]{article}
\usepackage{latex8}
\usepackage{times}
\usepackage{cite}
\usepackage{subfigure}
\usepackage{psfig}
\usepackage[pdftex]{graphicx}
\usepackage{epsfig}
\usepackage{pstricks}
\usepackage{fullpage}
\usepackage{amsthm}
\usepackage{graphics}
\usepackage{comment}
\usepackage[small,compact]{titlesec}
\usepackage[small,it]{caption}
\usepackage{wrapfig}

@@

\title{CSE530 Project Mid-Point Progress Report\\
Bandwidth-Aware Memory Hierarchy Design with Hybrid Memory Technologies}

@@

@@

\section{Project Description}

One of the challenges for modern chip-multiprocessor~(CMP) design is the growing bandwidth gap
between processor cores and off-chip main memory. Emerging memory technologies
bring in potential opportunities to fill this gap. The memory hierarchy,
however, need to be carefully designed in order to bridge the this gap with low
performance and power overhead.

In this project, we would like to explore the optimal memory hierarchy design
with various memory technologies specifically for bandwidth-bounded
applications. We will first estimate the performance of an application on a
baseline system, in terms of the bandwidth demanded by the application for a
range of memory capacities and the bandwidth that can be provided by the system
with such memory capacities. We would like to find out whether the bandwidth
provided by the memory hierarchy can satisfy the demand. If it is not satisfied,
we can modify the memory hierarchy, for example, by changing the capacity of
existing cache and memories, or by adding an extra level of cache or memory.
Based on our exploration, we would like to find a memory hierarchy design
solution optimized for bandwidth, including the number of cache and memory
levels, as well as the memory technology, the bandwidth, and the capacity used
for each level. In addition, we will enhance the memory hierarchy design from
the energy-efficiency point of view by constraining the total power consumption
of the memory system.\vspace{0.15in}

\section{Motivation}

@@

It is known that performance improvement of a computing system can be achieved
via multiple memory levels. Adding an extra level of memory, however, can also
help alleviate the bandwidth bottleneck of off-chip memory. In addition,
emerging memory technologies such as Magnetoresistive Random Access
Memory~(MRAM), Phase-Change Memory~(PCM), Resistive Random Access Memory~(RRAM),
etc., have shown potential to to be used as on-chip caches and the main
memory~\cite{Sun:2009:MRAM-L2-CMP}. Therefore, we would like to explore how to
enhance the memory hierarchy from the bandwidth point of view. Specifically, we
would like to examine (1) the number of levels in the optimal memory hierarchy,
(2) the appropriate memory technology of each level, and (3) the capacity and
bandwidth of each level. We will also explore the energy-efficiency of the
memory hierarchy design constrain, and the memory hierarchy design within a
fixed power budge.\vspace{0.15in}

@@

\section{Experimental Setup}

We will use our modified CACTI~\cite{CACTI} to estimate the bandwidths provided
by different memory technologies. We will also use
Simics~\cite{Magnusson:2002:simics-orig-paper} as the simulator to evaluate our
design method. The benchmarks will be selected from bandwidth-bound
applications, which possibly include benchmarks from SPEC
CPU2006~\cite{SPEC:2006} and NPB~\cite{NPB}.\vspace{0.15in}

@@

>>>>

\documentclass[10pt,letterpaper,singlecolumn]{article}
\usepackage{setspace}
\doublespacing
\usepackage{latex8}
\usepackage{times}
\usepackage{cite}
\usepackage{subfigure}
\usepackage{psfig}
\usepackage[pdftex]{graphicx}
\usepackage{epsfig}
\usepackage{pstricks}
\usepackage{fullpage}
\usepackage{amsthm}
\usepackage{graphics}
\usepackage{comment}
\usepackage[small,compact]{titlesec}
\usepackage[small,it]{caption}
\usepackage{wrapfig}

@@

\title{CSE530 Project Final Report\\
Bandwidth-Aware Memory Hierarchy Design with Hybrid Memory Technologies}

@@

\begin{abstract}
\boldmath

\end{abstract}

@@

\section{Introduction}

@@

It is known that performance improvement of a computing system can be achieved
via multiple memory levels. Adding an extra level of memory, however, can also
help alleviate the bandwidth bottleneck of off-chip memory. Specifically, we
would like to examine (1) the number of levels in the optimal memory hierarchy,
(2) the appropriate memory technology of each level, and (3) the capacity and
bandwidth of each level. We will also explore the energy-efficiency of the
memory hierarchy design constrain, and the memory hierarchy design within a
fixed power budge.\vspace{0.15in}

% -----------------------------------------------------------------------------
\section{Motivation}

Throughput computing involves performing a huge number of calculations with a large amount of parallelism. Throughput computing applications span many domains and are already critical on a variety of platforms from high-performance computing (HPC) machines to commercial servers to client machines.

Systems designed to perform well on throughput computing applications achieve high performance by exploiting their inherent parallelism. These systems support large numbers of threads and/or use wide SIMD execution. This puts a lot of pressure on the memory system. Fortunately, memory latency is typically not a bottleneck since the latency can be hidden via multithreading or hardware prefetching, since the data access patterns are relatively simple. On the other hand, bandwidth is a potential bottleneck.

Many throughput computing applications have inherently large working sets (i.e., tens to hundreds of MB). These are unlikely to fit in conventional on-die SRAM caches for the foreseeable future. Further, unlike more traditional workloads (e.g., those similar to SPEC or TPC benchmarks), many throughput computing applications see a sharp drop in performance once caches are too small to hold their working sets. Thus, these applications are likely to be bandwidth-bound at main memory unless some significant changes are made to the memory hierarchy.

Unfortunately, there are few simple techniques to improve bandwidth efficiency of a system. While there is some work in this area~\cite{BW-Wall}, this is insufficient for the large bandwidth requirements of some throughput computing applications. So today¡¯s systems that tout good performance for throughput computing applications (e.g., nVidia¡¯s Tesla) do so by providing large main memory bandwidths via the use of GDDR rather than improving bandwidth efficiency. However, GDDR has fairly strict capacity limits and is much more power hungry than conventional DRAM modules. This reduces its desirability for throughput computing, and makes it a bad choice for general purpose systems trying to improve their throughput computing performance.

Emerging memory technologies such as Magnetoresistive Random Access Memory~(MRAM), Phase-Change Memory~(PCM), Resistive Random Access Memory~(RRAM),
embedded DRAM, have shown potential to to be used as on-chip caches and the main memory~\cite{Sun:2009:MRAM-L2-CMP}. Therefore, we explore how to enhance the memory hierarchy from the bandwidth point of view.

@@

\section{Emerging Memory Technologies}

We first review the technology background three types of non-volatile memories, which are STT-RAM, PCRAM, and RRAM.

\subsection{STT-RAM}

\subsection{PCRAM}

\subsection{RRAM}

% -----------------------------------------------------------------------------
\section{Memory Technology Pre-Exploration}

We use our modified CACTI~\cite{CACTI} to estimate the bandwidths provided
by different memory technologies.

@@

 We will also use
Simics~\cite{Magnusson:2002:simics-orig-paper} as the simulator to evaluate our
design method. The benchmarks will be selected from bandwidth-bound
applications, which possibly include benchmarks from SPEC
CPU2006~\cite{SPEC:2006} and NPB~\cite{NPB}.\vspace{0.15in}

\subsection{Experimental Setup}

\subsection{Results}

\section{Conclusion and Future Work}
